Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-10-24 paper-new/concrete-relational-model.tex
revise intro; edit the two remarks in the relational model section

\begin{remark}
  The reason the set of perturbations needs to carry a monoid structure is that
  we will need to produce a syntactic perturbation corresponding to the identity
  relations $r(A)$, as well as one for the composition of relations. The monoid
  structure gives us what we need to produce these elements. Likewise, the
  reason that the interpretations as semantic perturbations must be
  \emph{homomorphisms} of monoids is that we need a syntactic perturbation that
  is interpreted as the identity semantic perturbation, and likewise given two
  syntactic perturbations $\delta_1$ and $\delta_2$ we need a syntactic
  pertubration $\delta_{1,2}$ that is interpreted as the composition of the
  semantic perturbations $i_A(\delta_1)$ and $i_A(\delta_2)$.
\end{remark}

\begin{remark}
  It might seem strange that in defining the morphisms of value and computation
  objects, there is no requirement involving the monoids. For example, one could
  imagine defining a morphism between value objects $(A, M_A, i_A)$ and $(A',
  M_{A'}, i_{A'})$ to be a predomain morphism $f \colon A \to A'$ along with a
  monoid homomorphism $h \colon M_A \to M_{A'}$ such that $f$ and $h$ satisfied
  a compatibility condition with respect to the interpretations $i_A$ and
  $i_{A'}$. However, with this definition of morphism, the category of value
  types is no longer Cartesian closed and is not a model of call-by-push-value.
  This is unfortunate, as it would simplify some aspects of the model construction.
\end{remark}

@@

Finally, we establish the core result of the semantics: interpreting
term precision as extensional squares. Fortunately, the hard work is
now complete, and the term precision rules follow from our
compositional constructions.
%
In particular, the congruence rules of term precision are modeled easily: congruence
squares exist both for bisimilarity and for the error ordering, and so
we also have these for extensional squares.
%
The \textsc{ErrBot} rule holds because $\mho$ is the least element in the
ordering on the free error domain. 
%
Next, the \textsc{EquivTyPrec} rule follows because $c \equiv c'$
implies quasi-equivalence of the denoted relations, and this in turn implies that
$\id \ltsqbisim{\sem{c}}{\sem{c'}} \id$.
%
Lastly, it remains to verify the existence of the squares for the cast rules. For $\upl$, the
relevant extensional square is obtained as follows:
%
% https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMSwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMCwzLCJlX2MiLDIseyJjdXJ2ZSI6MX1dLFsxLDQsIlxcZGVscmVfYyIsMl0sWzIsNSwiXFxwdXNoX3tjJ30oXFxkZWxyZV9jKSIsMix7ImN1cnZlIjoxfV0sWzAsMSwiYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEsMiwiYyciLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFszLDQsInIoQV8yKSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzQsNSwiYyciLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsyLDUsIlxcaWQiLDAseyJjdXJ2ZSI6LTF9XSxbMCwzLCJlX2MiLDAseyJjdXJ2ZSI6LTF9XSxbOCwxMywiXFxiaXNpbSIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs2LDE0LCJcXGJpc2ltIiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}[ampersand replacement=\&,column sep=large]
	{A_1} \& {A_2} \& {A_3} \\
	{A_2} \& {A_2} \& {A_3}
	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow[""{name=0, anchor=center, inner sep=0}, "{e_c}"', curve={height=6pt}, from=1-1, to=2-1]
	\arrow[""{name=1, anchor=center, inner sep=0}, "{e_c}", curve={height=-6pt}, from=1-1, to=2-1]
	\arrow["{c'}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
	\arrow["{\delre_c}"', from=1-2, to=2-2]
	\arrow[""{name=2, anchor=center, inner sep=0}, "{\push_{c'}(\delre_c)}"', curve={height=6pt}, from=1-3, to=2-3]
	\arrow[""{name=3, anchor=center, inner sep=0}, "\id", curve={height=-6pt}, from=1-3, to=2-3]
	\arrow["{r(A_2)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
	\arrow["{c'}"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
	\arrow["\bisim"{description}, draw=none, from=0, to=1]
	\arrow["\bisim"{description}, draw=none, from=2, to=3]
\end{tikzcd}\]
%
% Note that the horizontal composition of the squares is defined, as they are
% intensional squares.
We have used the push-pull structure on $c'$ to turn the perturbation on $A_2$
into a perturbation on $A_3$. Lastly we note that the bottom edge of the square
is equal to $c'$ by the fact that $c'$ is downward closed. The squares for $\upr$,
$\dnl$, and $\dnr$ are obtained analogously.

>>>>

\begin{remark}
  The reason why the set of syntactic perturbations needs to carry a monoid
  structure and why the interpretation needs to be a homomorphism of monoids is
  related to the construction of the identity relations and composition of
  relations on value/computation types. As part of this construction, we will
  need to produce a syntactic perturbation corresponding to the identity
  predomain/error domain relation, as well as corresponding to the composition
  of relations. The monoid structure gives us what we need to produce these
  perturbations.
  % we will need to produce a syntactic perturbation corresponding to the identity
  % relations $r(A)$, as well as one for the composition of relations. The monoid
  % structure gives us what we need to produce these elements. Likewise, the
  % reason that the interpretations as semantic perturbations must be
  % \emph{homomorphisms} of monoids is that we need a syntactic perturbation that
  % is interpreted as the identity semantic perturbation, and likewise given two
  % syntactic perturbations $\delta_1$ and $\delta_2$ we need a syntactic
  % pertubration $\delta_{1,2}$ that is interpreted as the composition of the
  % semantic perturbations $i_A(\delta_1)$ and $i_A(\delta_2)$.
  % These properties are used to define the identity and composition of value and computation relations
\end{remark}

% It doesn't work: not all types have such an action, e.g., the function type, because of the definition of the perturbations
% The fact that we cannot define the morphisms this way is unfortunate, since doing so would simplify
% some aspects of the model construction.
\begin{remark}
  It might seem strange that in defining the morphisms of value and computation
  objects, there is no requirement involving the monoids. For example, one could
  imagine defining a morphism between value objects $(A, M_A, i_A)$ and $(A',
  M_{A'}, i_{A'})$ to be a predomain morphism $f \colon A \to A'$ along with a
  monoid homomorphism $h \colon M_A \to M_{A'}$ such that $f$ and $h$ satisfy a
  compatibility condition with respect to the interpretations $i_A$ and
  $i_{A'}$. However, with this definition we cannot interpret the function type,
  since the morphisms defined in this way do not themselves form a value type. 
  % This is due to the way that the syntactic perturbations are defined for the function type.
  %
  The fact that we cannot define the morphisms in this way is unfortunate, since
  doing so would simplify some aspects of the model construction.

  % However, with this definition of morphism, the category of value
  % types is no longer Cartesian closed and is not a model of call-by-push-value.
  % This is unfortunate, as it would simplify some aspects of the model construction.
\end{remark}

@@

Finally, we establish the core result of the semantics: interpreting
term precision as extensional squares. Fortunately, the hard work is
now complete, and the term precision rules follow from our
compositional constructions.
%
In particular, the congruence rules of term precision are modeled easily: congruence
squares exist both for bisimilarity and for the error ordering, and so
we also have these for extensional squares.
%
The \textsc{ErrBot} rule holds because $\mho$ is the least element in the semantics of the . 
%
Next, the \textsc{EquivTyPrec} rule follows because $c \equiv c'$
implies quasi-equivalence of the denoted relations, and this in turn implies that
$\id \ltsqbisim{\sem{c}}{\sem{c'}} \id$.
%
Lastly, it remains to verify the existence of the squares for the cast rules. For $\upl$, the
relevant extensional square is obtained as follows:
%
% https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMSwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMCwzLCJlX2MiLDIseyJjdXJ2ZSI6MX1dLFsxLDQsIlxcZGVscmVfYyIsMl0sWzIsNSwiXFxwdXNoX3tjJ30oXFxkZWxyZV9jKSIsMix7ImN1cnZlIjoxfV0sWzAsMSwiYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEsMiwiYyciLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFszLDQsInIoQV8yKSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzQsNSwiYyciLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsyLDUsIlxcaWQiLDAseyJjdXJ2ZSI6LTF9XSxbMCwzLCJlX2MiLDAseyJjdXJ2ZSI6LTF9XSxbOCwxMywiXFxiaXNpbSIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs2LDE0LCJcXGJpc2ltIiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}[ampersand replacement=\&,column sep=large]
	{A_1} \& {A_2} \& {A_3} \\
	{A_2} \& {A_2} \& {A_3}
	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow[""{name=0, anchor=center, inner sep=0}, "{e_c}"', curve={height=6pt}, from=1-1, to=2-1]
	\arrow[""{name=1, anchor=center, inner sep=0}, "{e_c}", curve={height=-6pt}, from=1-1, to=2-1]
	\arrow["{c'}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
	\arrow["{\delre_c}"', from=1-2, to=2-2]
	\arrow[""{name=2, anchor=center, inner sep=0}, "{\push_{c'}(\delre_c)}"', curve={height=6pt}, from=1-3, to=2-3]
	\arrow[""{name=3, anchor=center, inner sep=0}, "\id", curve={height=-6pt}, from=1-3, to=2-3]
	\arrow["{r(A_2)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
	\arrow["{c'}"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
	\arrow["\bisim"{description}, draw=none, from=0, to=1]
	\arrow["\bisim"{description}, draw=none, from=2, to=3]
\end{tikzcd}\]
%
% Note that the horizontal composition of the squares is defined, as they are
% intensional squares.
We have used the push-pull structure on $c'$ to turn the perturbation on $A_2$
into a perturbation on $A_3$. Lastly we note that the bottom edge of the square
is equal to $c'$ by the fact that $c'$ is downward closed. The squares for $\upr$,
$\dnl$, and $\dnr$ are obtained analogously.

### 2024-10-24 paper-new/intro.tex
revise intro; edit the two remarks in the relational model section

The only prior denotational semantics for gradual typing was given by
New and Licata and is based on a classical Scott-style \emph{domain
theory} \cite{new-licata18}. The fundamental idea is to equip
$\omega$-CPOs with an additional ``error ordering'' $\ltdyn$ which
models the graduality ordering, and for casts to arise from
\emph{embedding-projection pairs}. Then the graduality property
follows as long as all language constructs can be interpreted using
constructions that are monotone with respect to the error ordering.
%
This framework has the benefit of being compositional, and was
expressive enough to be extended to model dependently typed gradual
typing \cite{gradualizing-cic}.
%
However, an approach based on classical domain theory has fundamental
limitations: domain theory is incapable of modeling certain perversely
recursive features of programming languages such as dynamic type tag
generation and higher-order references, which are commonplace in
real-world gradually typed systems as well as gradual calculi.
%
Our long-term goal is to develop a denotational approach that can
scale up to these advanced features, and so we must abandon classical
domain theory as the foundation in order to make progress. In this
work, we provide first steps towards this goal by adapting work on a
simple gradually typed lambda calculus to \emph{guarded domain
theory}. Our goal is for this model to scale to dynamic type tag
generation and higher-order references in future work.

@@

% The theorems we want are about the whole sequence of approximations
% whereas working with presheaves is about each finite approximation
% The statement of graduality is in terms of global elements
\eric{This next paragraph could be moved to the section on adequacy, but since
the example is in the analytic setting it also makes sense to keep it here. I'm
not sure which option makes more sense.} While the definitions in guarded domain
theory are constructed as sequences of approximations, results about semantics
of programs, e.g., graduality, are actually statements concerning entire
\emph{sequences} of approximations. As a simple example, consider the set of
programs that may take a computational step or return unit.\footnote{This
example is adapted from \cite{mogelberg-paviotti2016}.} In the topos of trees,
we can model the set of these programs as the indexed family of sets $\{X_n\}_{n
\in \mathbb{N}}$ where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1,
\bot\}$. Here, $i$ denotes a program that steps $i$ times and then returns, and
$\bot$ denotes a program that fails to terminate in $n-1$ steps or fewer. For
any fixed $n$, the set $X_n$ fails to distinguish a program that takes $n$ steps
and then terminates from a program that never terminates. On the other hand, if
we define the denotation of a program to be a \emph{sequence} of elements $x_i
\in X_i$ for all $i \in \mathbb{N}$, then the denotation of the diverging

%% \max{what is the point you are trying to convey in this paragraph?}
%% Because the solutions we obtain working in guarded domain theory are constructed
%% as a sequence of increasingly better approximations, when we want to establish a
%% property that holds in the limit, we must reason about all finite
%% approximations. In the setting of the topos of trees, this manifests as the need
%% to consider \emph{global elements} of the presheaves, i.e., a family of elements
%% $x_i \in X_i$ compatible with the restriction maps. For example, consider the
%% set of programs that may take a computational step or return unit.\footnote{This example is adapted from \cite{mogelberg-paviotti2016}.} In the
%% topos of trees, we can model this set as the indexed family of sets $\{X_n\}_{n \in \mathbb{N}}$
%% where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1, \bot\}$. Here, $i$ denotes a program that steps
%% $i$ times and then returns, and $\bot$ denotes a program that fails to terminate
%% in $n-1$ steps or fewer. For any fixed $n$, the set $X_n$ fails to distinguish a
%% program that takes $n$ steps and then terminates from a program that never
%% terminates. On the other hand, if we instead consider the denotation of a
%% program to be a global element of the presheaf $\{X_n\}_{n \in \mathbb{N}}$ -- a family of elements
%% $x_i \in X_i$ for all $i \in \mathbb{N}$ -- then the denotation of the diverging
%% >>>>>>> Stashed changes
program will be distinct from that of a program that terminates, regardless of
the number of steps it takes.

% Because the solutions we obtain working in guarded domain theory are constructed
% as a sequence of increasingly better approximations, when we want to establish a
% property that holds in the limit, we must reason about all finite
% approximations. In the setting of the topos of trees, this manifests as the need
% to consider \emph{global elements} of the presheaves, i.e., a family of elements
% $x_i \in X_i$ compatible with the restriction maps.

% \max{this next paragraph is a bit too mysterious}
% This has important ramifications for defining an adequate denotational semantics
% in guarded domain theory, as we seek to do in this paper: we do not want our
% denotational semantics to conflate a diverging program with a program that fails
% to terminate in sufficiently few steps. Thus, establishing adequacy of a
% semantics in guarded domain theory will require a means of globalizing,
% a point we will return to in Section \ref{sec:big-step-term-semantics}.

@@

While guarded domain theory can be presented analytically using
ultrametric spaces or the topos of trees, in practice it is
considerably simpler to work \emph{synthetically} by working in a
non-standard foundational system such as \emph{guarded type
theory}. In guarded type theory later is taken as a primitive
operation on types, and we take as an axiom that guarded domain
equations have a (necessarily unique) solution. The benefit of this
synthetic approach is that when working in the non-standard
foundation, we don't need to model an object language type as a
step-indexed set, but instead simply as a set, and object-language
terms can be modeled in the Kleisli category of a simple monad defined
using guarded recursion. Not only does this make on-paper reasoning
about guarded domain theory easier, it also enables a simpler avenue
to verification in a proof assistant. Whereas formalizing analytic
guarded domain theory would require a significant theory of presheaves
and making sure that all constructions are functors on categories of
presheaves, formalizing synthetic guarded domain theory can be done by
directly adding the later modality and the guarded fixed point
property axiomatically.
%
% \max{TODO: more here, specifically this is where we should talk about adequacy I think}
%
In this paper, we work informally in a guarded type theory which is described in
more detail in Section \ref{sec:guarded-type-theory}.
%
% As in the setting of analytic guarded domain theory, in the synthetic setting it
% is also possible to construct global solutions. The specific approach we use in
% this work is that of \emph{clocks} and \emph{clock-quantification}
% \cite{atkey-mcbride2013}.

@@

Our approach to proving graduality by constructing a denotational model in
guarded type theory consists of two main steps. First, there is the construction
of the model itself. This gives us an interpretation of the terms as well as the
axioms that model the graduality property. However, since this model is defined
in guarded domain theory, the denotations of terms are constructed as sequences of
approximations and our graduality result is actually a statement about each
finite approximation. The actual graduality property should be a statement about
entire sequences of approximations.
%
What we need is to show that the guarded model induces a sensible set-theoretic
semantics in ordinary mathematics. We cannot hope to derive such a semantics for
\emph{all} types (e.g., the dynamic type), but for the subset of closed terms of
base type, we should be able to extract a well-behaved semantics for which the
graduality property holds.

@@

To define the big-step semantics, we apply the technique of
\emph{clock-quantification} \cite{atkey-mcbride2013}, which is the synthetic
analogue of computing global elements in analytic guarded domain theory. This
gives us an interpretation of closed programs of base type as partial functions.
We describe the process in more detail in Section
\ref{sec:big-step-term-semantics}. The resulting term semantics is adequate, and
furthemore it validates the equational theory, since equivalent terms in the
equational theory denote equal terms in the big-step semantics.

@@

The main contribution of this work is a compositional denotational semantics in
guarded type theory for a simple gradually typed language that validates
$\beta\eta$ equality and satisfies a graduality theorem.
%
% In particular, the notion of syntactic perturbations stands out as a key
% technical contribution and is the most significant and novel feature of our
% model that allowed us to successfully adapt the denotational approach of New and
% Licata to the guarded setting.
% In particular, the notion of syntactic perturbations stands out as a key feature
% of our model, and is our most significant technical contribution allowing us to
% successfully adapt the denotational approach of New and Licata to the guarded
% setting.
The goal we have kept in mind while designing our model is to handle
gradually-typed languages with advanced features that classical domain theory in
incapable of modeling. We do not address these features in this work but leave
them instead to immediate future work, as handling even the simplest
gradually-typed language already poses interesting challenges.
%
Within our semantics, the notion of syntactic perturbation is our most
significant technical contribution, allowing us to successfully adapt the
denotational approach of New and Licata to the guarded setting.
%
Most of our work has further been verified in Guarded Cubical Agda
\cite{veltri-vezzosi2020}, demonstrating that the semantics is readily
mechanizable. We provide an overview of the mechanization effort, and what
remains to be formalized, in Section \ref{sec:mechanization}.
%

@@

The paper is laid out as follows:
% \max{I think there's too much detail in this outline, especially parts 4 and 5. Can we instead expand those into the Intro?}
\begin{enumerate}
\item In Section \ref{sec:GTLC} we fix our input language, a fairly
  typical gradually typed cast calculus.
\item In Section \ref{sec:concrete-term-model} we develop a
  denotational semantics in synthetic guarded domain theory for the
  \emph{terms} of the gradual lambda calculus.  The model is adeqauate
  and validates the equational theory, but it does not satisfy
  graduality. We use this to introduce some of our main technical
  tools: modeling recursive types in guarded type theory and modeling
  effects using call-by-push-value.
\item In Section \ref{sec:towards-relational-model} we show where the New-Licata
  classical domain theoretic approach fails to adapt cleanly to the guarded
  setting and explore the difficulties of proving graduality in an intensional
  model. We prove the no-go theorem about extensional, transitive relations,
  introduce the lock-step error ordering and weak bisimilarity relation, and
  motivate the need for perturbations.
\item In Section \ref{sec:concrete-relational-model} we describe the
  construction of the model in detail, and discuss the
  translation of the syntax and axioms of the gradually typed cast calculus into the model.
  Lastly, we prove that the model is adequate for the graduality property.
  % The resulting model validates the axioms for type and term precision specified
  % in Section \ref{sec:GTLC}. The final step is to prove that the model is
  % \emph{adequate} for the graduality property: a closed term precision $M \ltdyn
  % N : \nat$ has the expected semantics, i.e., that $M$ errors or $M$ and $N$
  % have the same extensional behavior. We do so by extending the globalization
  % techniques used in defining the big-step term model to account for the
  % interpretation of term precision.
\item In Section \ref{sec:discussion} we discuss prior work on proving
  graduality, the partial mechanization of our results in Agda, and
  future directions for denotational semantics of gradual typing.
\end{enumerate}

>>>>

The only prior denotational semantics for gradual typing was given by
New and Licata and is based on a classical Scott-style \emph{domain
theory} \cite{new-licata18}. The fundamental idea is to equip
$\omega$-CPOs with an additional ``error ordering'' $\ltdyn$ which
models the graduality ordering, and for casts to arise from
\emph{embedding-projection pairs}. Then the graduality property
follows as long as all language constructs can be interpreted using
constructions that are monotone with respect to the error ordering.
%
This framework has the benefit of being compositional, and was
expressive enough to be extended to model dependently typed gradual
typing \cite{gradualizing-cic}.
%
However, an approach based on classical domain theory has fundamental
limitations: domain theory appears incapable of modeling certain perversely
recursive features of programming languages such as dynamic type tag
generation and higher-order references \cite{Birkedal-Stovring-Thamsborg-2009}, which are commonplace in
real-world gradually typed systems as well as gradual calculi.
%
Our long-term goal is to develop a denotational approach that can
scale up to these advanced features, and so we must abandon classical
domain theory as the foundation in order to make progress. In this
work, we provide first steps towards this goal by adapting work on a
simple gradually typed lambda calculus to \emph{guarded domain
theory}. Our goal is for this model to scale to dynamic type tag
generation and higher-order references in future work.

@@

% The theorems we want are about the whole sequence of approximations
% whereas working with presheaves is about each finite approximation
% The statement of graduality is in terms of global elements
% \eric{This next paragraph could be moved to the section on adequacy, but since
% the example is in the analytic setting it also makes sense to keep it here. I'm
% not sure which option makes more sense.} While the definitions in guarded domain
% theory are constructed as sequences of approximations, results about semantics
% of programs, e.g., graduality, are actually statements concerning entire
% \emph{sequences} of approximations. As a simple example, consider the set of
% programs that may take a computational step or return unit.\footnote{This
% example is adapted from \cite{mogelberg-paviotti2016}.} In the topos of trees,
% we can model the set of these programs as the indexed family of sets $\{X_n\}_{n
% \in \mathbb{N}}$ where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1,
% \bot\}$. Here, $i$ denotes a program that steps $i$ times and then returns, and
% $\bot$ denotes a program that fails to terminate in $n-1$ steps or fewer. For
% any fixed $n$, the set $X_n$ fails to distinguish a program that takes $n$ steps
% and then terminates from a program that never terminates. On the other hand, if
% we define the denotation of a program to be a \emph{sequence} of elements $x_i
% \in X_i$ for all $i \in \mathbb{N}$, then the denotation of the diverging
% program will be distinct from that of a program that terminates, regardless of
% the number of steps it takes.

@@

While guarded domain theory can be presented analytically using
ultrametric spaces or the topos of trees, in practice it is
considerably simpler to work \emph{synthetically} by working in a
non-standard foundational system such as \emph{guarded type
theory}. In guarded type theory later is taken as a primitive
operation on types, and we take as an axiom that guarded domain
equations have a (necessarily unique) solution. The benefit of this
synthetic approach is that when working in the non-standard
foundation, we don't need to model an object language type as a
step-indexed set, but instead simply as a set, and object-language
terms can be modeled in the Kleisli category of a simple monad defined
using guarded recursion. Not only does this make on-paper reasoning
about guarded domain theory easier, it also enables a simpler avenue
to verification in a proof assistant. Whereas formalizing analytic
guarded domain theory would require a significant theory of presheaves
and making sure that all constructions are functors on categories of
presheaves, formalizing synthetic guarded domain theory can be done by
directly adding the later modality and the guarded fixed point
property axiomatically.
%
% \max{TODO: more here, specifically this is where we should talk about adequacy I think}
%
In this paper, we work informally in a guarded type theory which is described in
more detail in Section \ref{sec:guarded-type-theory}.

% The move to synthetic setting is useful, but now you're stuck in the guarded setting
The simplicity afforded by working in the synthetic setting comes with a
drawback: by changing our ambient foundations so that all types represent
step-indexed sets, it is no longer possible to define a type that denotes an
ordinary set. Roughly speaking, what is needed is an ``escape hatch'' to leave
the guarded setting and make definitions that are interpreted as constructions
in ordinary mathematics. This is typically accomplished by adding a modality to
the guarded type theory. The specific approach we take in this paper is the
technique of \emph{clock-quantification} \cite{atkey-mcbride2013}. In
particular, the guarded type theory in which we work has a notion of clocks that
index all of our guarded constructions. This is discussed in more detail in
Section \ref{sec:big-step-term-semantics}.

%
% As in the setting of analytic guarded domain theory, in the synthetic setting it
% is also possible to construct global solutions. The specific approach we use in
% this work is that of \emph{clocks} and \emph{clock-quantification}
% \cite{atkey-mcbride2013}.

@@

% Our approach to proving graduality by constructing a denotational model in
% guarded type theory consists of two main steps. First, there is the construction
% of the model itself. This gives us an interpretation of the terms as well as the
% axioms that model the graduality property.

Having defined the model of gradual typing in guarded type theory, we need to
show that the guarded model induces a sensible set-theoretic semantics in
ordinary mathematics. We cannot hope to derive such a semantics for \emph{all}
types (e.g., the dynamic type), but for the subset of closed terms of base type,
we should be able to extract a well-behaved semantics for which the graduality
property holds.

@@

To define the big-step semantics, we apply clock quantification to obtain
an interpretation of closed programs of base type as partial functions.
We describe the process in more detail in Section
\ref{sec:big-step-term-semantics}. The resulting term semantics is adequate, and
furthemore it validates the equational theory, since equivalent terms in the
equational theory denote equal terms in the big-step semantics.

@@

The main contribution of this work is a compositional denotational semantics in
guarded type theory for a simple gradually typed language that validates
$\beta\eta$ equality and satisfies a graduality theorem.
%
% In particular, the notion of syntactic perturbations stands out as a key
% technical contribution and is the most significant and novel feature of our
% model that allowed us to successfully adapt the denotational approach of New and
% Licata to the guarded setting.
% In particular, the notion of syntactic perturbations stands out as a key feature
% of our model, and is our most significant technical contribution allowing us to
% successfully adapt the denotational approach of New and Licata to the guarded
% setting.
%
Within our semantics, the notion of syntactic perturbation is our most
significant technical contribution, allowing us to successfully adapt the
denotational approach of New and Licata to the guarded setting.
%
Most of our work has further been verified in Guarded Cubical Agda
\cite{veltri-vezzosi2020}, demonstrating that the semantics is readily
mechanizable. We provide an overview of the mechanization effort, and what
remains to be formalized, in Section \ref{sec:mechanization}.
%

@@

The paper is laid out as follows:
% \max{I think there's too much detail in this outline, especially parts 4 and 5. Can we instead expand those into the Intro?}
\begin{enumerate}
\item In Section \ref{sec:GTLC} we fix our input language, a fairly
  typical gradually typed cast calculus.
\item In Section \ref{sec:concrete-term-model} we develop a
  denotational semantics in synthetic guarded domain theory for the
  \emph{terms} of the gradual lambda calculus.  The model is adeqauate
  and validates the equational theory, but it does not satisfy
  graduality. We use this to introduce some of our main technical
  tools: modeling recursive types in guarded type theory and modeling
  effects using call-by-push-value.
\item In Section \ref{sec:towards-relational-model} we show where the New-Licata
  classical domain theoretic approach fails to adapt cleanly to the guarded
  setting and explore the difficulties of proving graduality in an intensional
  model. We prove a no-go theorem about extensional, transitive relations,
  introduce the lock-step error ordering and weak bisimilarity relation, and
  motivate the need for perturbations.
\item In Section \ref{sec:concrete-relational-model} we describe the
  construction of the model in detail, and discuss the
  translation of the syntax and axioms of the gradually typed cast calculus into the model.
  Lastly, we prove that the model is adequate for the graduality property.
  % The resulting model validates the axioms for type and term precision specified
  % in Section \ref{sec:GTLC}. The final step is to prove that the model is
  % \emph{adequate} for the graduality property: a closed term precision $M \ltdyn
  % N : \nat$ has the expected semantics, i.e., that $M$ errors or $M$ and $N$
  % have the same extensional behavior. We do so by extending the globalization
  % techniques used in defining the big-step term model to account for the
  % interpretation of term precision.
\item In Section \ref{sec:discussion} we discuss prior work on proving
  graduality, the partial mechanization of our results in Agda, and
  future directions for denotational semantics of gradual typing.
\end{enumerate}
